---
title: "Exercise 2"
author: "Stefano D'Arrigo & Martina Betti"
date: "5/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
library(tensorflow)
```

## Look into the data

Load the data

```{r}
load("../data/amazon_review_clean.RData")
```

The X–matrices are known as term-frequency matrices or document-term matrices and count how often various words are used in the text. For more info have also a look at the Kernel for Text slide set (pp. 142-147).
Specifically, the X[i,j] element counts the number of times the word wj appears in the ith review for some pre-specified set of words of interest. Let's check:

```{r}
dim(X_tr)
colnames(X_tr)[1:10]
round(as.matrix(X_tr[1:10, 1:10]), 2)
```


Implement a linear classifier: 

```{r}
# Define the regression function
f_reg <- function(X, beta) tf$matmul(cbind(1,X), beta)

# Define the linear classifier
linear_classifier <- function(X, beta) ifelse(f_reg(X, beta) >= 0.5, 1, 0)

# Define the gradient for the linear
gradient_linear <- function(X.T_X, X.T_Y, beta) tf$matmul(X.T_X, beta) - X.T_Y

# Define the empirical risk
empirical_risk <- function(y.pred.reg, y.true) mean(as.matrix((y.pred.reg-y.true)^2)) # OLS

# Define the gradient descent function
gradient_descent <- function(X, Y, G, alpha=0.5, max_iter=100, tol=0.00001, beta=NULL, plot=FALSE, threads = 4) {
  ###
  # Function to compute the gradient descent given the gradient function G.
  # :param X : design matrix (n_observations * n_features)
  # :param Y : ground-truth labels
  # :param G : gradient function
  # :param alpha : learning rate
  # :param max_iter : maximum number of iterations; first stop condition
  # :param tol : tolerance; stop the gradient loop before max_iter is reached if the condition is met
  # :param beta : vector of parameters; provide a custom starting vector; otherwise, a random one is chosen
  # :param plot : boolean; plot the empirical risk history if TRUE 
  # :param threads : integer that defines the number of cores to use for the matrix precomputation
  #
  # :return list(beta=[vector of parameters], emp_risk_hist=[empirical risk through the iterations]) 
  ###
  # gradient = 2/n (X.T*X*beta - X.T*Y)
  
  n <- dim(X)[1] # number of observations (data points)
  d <- dim(X)[2] # number of features

  
  if(is.null(beta))
    beta <- matrix(runif(d+1, min = -1, max = 1))
  
  emp_risk_hist <- rep(0, max_iter)
  
  print("Precomputing the matrices")
  
  X.T_X = tf$matmul(t(cbind(1,X)), cbind(1,X)) * (2 / n) # (d+1) x (d+1)
  X.T_Y = tf$matmul(t(cbind(1,X)), Y) * (2 / n) # (d+1) x 1
  
  print("Entering gradient loop...")  
  
  # Gradient loop
  for(i in 1:max_iter) {

    grad <- G(X.T_X, X.T_Y, beta)
    
    # Update beta
    beta <- beta - alpha * grad # update rule

    # Compute and store the ER value of the i-th iteration
    emp_risk_hist[i] <- empirical_risk(f_reg(X, beta), Y)
    
    print(paste("Iteration", i, " ER:", emp_risk_hist[i]))
    
    # Check convergence
    if(all(abs(grad) <= tol)) {
      max_iter <- i
      break
    }
    
  }
  print(paste("Converged after", max_iter, "iterations"))

  if(plot)
    plot(seq(1,max_iter), emp_risk_hist[1:max_iter], col="blue",
         xlab="iter", ylab="ER", main="Gradient descent")

  return(list(param=beta, emp_risk_hist=emp_risk_hist))
}

```



Map the Y string labels into \{0,1\}:

```{r}
encode <- function(Y) {
  unique_labs <- unique(Y)
  Y_encoded <- as.integer(Y == unique_labs[2])
  return(matrix(Y_encoded))
}
```

Take a small random subset of the train set:

```{r eval=FALSE, include=FALSE}
n <- length(y_tr)
idxs <- sample(seq(1, n), size = n, replace = FALSE)
```

Encode the labels and compute the parameters of the trained model:

```{r}
y_tr_encoded <- encode(y_tr)# encode(y_tr[idxs])

t_0 <- Sys.time()
t_0

# with lr=0.003 the ER diverges, with lr=0.001 the ER converges, but slowly
result <- gradient_descent(X_tr, y_tr_encoded, gradient_linear, alpha=0.001, max_iter=500, plot=F, threads=8)
Sys.time() - t_0
```

Take a small random subset of the test set:

```{r}
n <- length(y_te)
idxs <- sample(seq(1, n), size = 1000, replace = FALSE)
```

Encode the labels of the ground-truth vector and compute the predictions:

```{r}
y_te_encoded <- encode(y_te[idxs])

pred <- linear_classifier(X_te[idxs,], result$param)
```

Rapidly compute the number of right predictions:

```{r}
perf <- sum(pred == y_te_encoded) / 1000
```

Check the number of right predictions:

```{r}
perf
```
```{r}
plot(seq(1, length(result$emp_risk_hist)), result$emp_risk_hist)
```

