---
title: "Exercise 2"
author: "Stefano D'Arrigo & Martina Betti"
date: "5/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Look into the data

Load the data

```{r}
load("../data/amazon_review_clean.RData")
```

The X–matrices are known as term-frequency matrices or document-term matrices and count how often various words are used in the text. For more info have also a look at the Kernel for Text slide set (pp. 142-147).
Specifically, the X[i,j] element counts the number of times the word wj appears in the ith review for some pre-specified set of words of interest. Let's check:

```{r}
dim(X_tr)
colnames(X_tr)[1:10]
round(as.matrix(X_tr[1:10, 1:10]), 2)
```


Implement a linear classifier: for simplicity, I've implemented logistic regression, but another one can be used.


```{r}
# Define the sigmoid function (logistic function)
sigmoid <- function(X, beta) 1 / (1 + exp(- t(X) %*% beta))


# Define the gradient for the logistic regression
gradient_logistic <- function(beta, X, XY) {
  n <- dim(X)[1] # number of observations (data points)
  retrun((t(X) %*% sigmoid(X, beta) - XY ) * 2 / n)
}


# Define the gradient descent function
gradient_descent <- function(X, Y, G, alpha=0.01, max_iter=100, tol=0.001, beta=NULL) {
  ###
  # Function to compute the gradient descent given the gradient function G.
  # :param X : design matrix (n_observations * n_features)
  # :param Y : ground-truth labels
  # :param G : gradient function
  # :param alpha : learning rate
  # :param max_iter : maximum number of iterations; first stop condition
  # :param tol : tolerance; stop the gradient loop before max_iter is reached if the condition is met
  #
  # :return beta [vector of coefficients] 
  ###
  
  # Setup
  if(is.null(beta))
    beta <- matrix(runif(d))
  
  n <- dim(X)[1] # number of observations (data points)
  d <- dim(X)[2] # number of features
  XY <- t(X) %*% Y # precompute the product between the design matrix X and the vector Y
  
  # Gradient loop
  for(i in 1:max_iter) {
    
    # Store the value of beta of the previous iteration
    beta_old <- beta 
    
    # Update beta according to the gradient descent rule
    beta <- beta - alpha * G(beta, X, XY)
    
    # Check convergence 
    if(all(abs(beta - beta_old) <= tol)) {
      print(paste("Converged after", i, "iterations"))
      return(beta)
    }
  }
  print(paste("Maximum number of iterations reached: ", max_iter))
  return(beta)
}
```

Map the Y string labels into \{0,1\}.

```{r}
encode <- function(Y) {
  unique_labs <- unique(Y)
  Y_encoded <- sapply(Y, function(y) {
    if(y == unique_labs[1]) return(0)
    else return(1)
  })
  return(Y_encoded)
}
```



```{r}
y_encoded <- encode(y_tr)

beta <- gradient_descent(X_tr, y_encoded, gradient_logistic, alpha=0.5)
```


